Mainstreaming Climate Adaptation Requires Subjective Integration of Deep Uncertainties

### 2022-05-24 papers/paper/preprint.tex
Updates from Overleaf

\documentclass[11pt]{article}

% Figures
\usepackage{graphicx}
\usepackage[list=true]{subcaption}
\graphicspath{{../../plots/}{../../tikz/}{../../img/}}
\usepackage[section]{placeins} % require floats to appear in the section they are defined

@@

% footer
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage[en-US]{datetime2}
\fancyhf{}
\fancyhead[L]{CONFIDENTIAL DRAFT by J. Doss-Gollin \& K. Keller}
\fancyhead[R]{\DTMnow}
\fancyfoot[R]{page~\thepage~of~\pageref{LastPage}}
\pagestyle{fancy}
\setlength{\headheight}{15pt}

@@

% authors
\usepackage{authblk}
\title{A subjective Bayesian framework for synthesizing deep uncertainties in climate risk management}
\author[1]{James Doss-Gollin}
\author[2]{Klaus Keller}
\affil[1]{Department of Civil and Environmental Engineering, Rice University}
\affil[2]{Thayer School of Engineering, Dartmouth College}
\renewcommand*{\Affilfont}{\normalsize\normalfont}

% ACRONYMS
\usepackage[acronym, nopostdot, nonumberlist, shortcuts, numberedsection, nogroupskip,]{glossaries}
\newacronym{bfe}{BFE}{base flood elevation}
\newacronym{cmip}{CMIP}{the Coupled Model Intercomparison Project}
\newacronym{dmdu}{DMDU}{decision making under deep uncertainty}
\newacronym{fema}{FEMA}{the Federal Emergency Management Agency}
\newacronym{gev}{GEV}{generalized extreme value}
\newacronym{iid}{IID}{independent and identically distributed}
\newacronym{ipcc}{IPCC}{International Panel on Climate Change}
\newacronym{msl}{MSL}{mean relative sea level}
\newacronym{noaa}{NOAA}{the National Oceanic and Atmospheric Administration}
\newacronym{pdf}{PDF}{probability density function}
\newacronym{rcp}{RCP}{representative concentration pathway}
\newacronym{slr}{SLR}{sea level rise}
\newacronym{ssp}{SSP}{shared socio-economic pathway}
\newacronym[]{usace}{USACE}{United States Army Corps of Engineers}
\newacronym[]{usgs}{USGS}{United States Geological Survey}
\newacronym[plural=SOWs,firstplural=states of the world (SOWs),descriptionplural=states of the world]{sow}{SOW}{state of the world}
\makeglossaries % required to make the list of acronyms

@@

\begin{abstract}
    Methods for doing so, whether classical standards-based approaches or bottom-up robust methods, make implicit or explicit assumptions about the likelihood of different futures.
    These assumptions may be inconsistent with expert belief.
    In this paper we propose a probabilsitic method for synthesizing across deep uncertainties, using subjective prior beliefs.
    The proposed method can be used in tandem with bottom-up exploratory methods and is consistent with fully probabilistic design criteria.
    Through the lens of a didactic case study of house elevation in Norfolk, VA, we illustrate how this approach can improve the transparency of modeling assumptions.\klaus{Abstract needs a little work}
\end{abstract}

Key points
\begin{enumerate}
    \item Estimated trade-offs between adaptation objectives are sensitive to assumptions about the likelihood of future conditions.
    \item We propose a probabilistic method for transparently synthesizing across deep uncertainty in future conditions.
    \item The proposed method identifies strategies for house elevation that are robust to assumptions in sea level rise.
\end{enumerate}

@@

Critical infrastructure services in America are aging and inadequate, leading to high-impact failures even during conditions with substantial precedent in the historical record \citep[\eg,][]{doss-gollin_txtreme:2021,doss-gollin_fatalism:2020}.
Looking to the future, it is expected that changes to regulation, economics and finance, patterns of population and infrastructure use, and climate will further stress infrastructure systems.
A critical question, then, is what conditions or futures particular infrastructure systems and components should be designed to withstand.
The answer to this question depends in part on values: intrinsic trade-offs between safety, performance, cost, and other objectives depend on context and stakeholder preferences \citep{keller_management:2021}.
However, this decision is also subject to assumptions regarding the likelihhod and relevance of possible futures.

\subsection{Current practice}

Current practice in engineering, infrastructure design, and local governance relies heavily upon standards that specify particular ``design events'' that buildings and infrastructure should safely withstand \citep{asce_7-10:2013,bruneau_multihazard:2017}.
These are often, though not always, based on probabilistic analysis of relevant data.
For example, \gls{fema}, local governments, and engineering consultants produce local floodplain maps in many communities.
These trigger specific floodplain regulations, such as a nominal requirement for homes purchased through using federally backed mortgage porgrams to be covered by flood insurance \citep{kousky_voucher:2014}.
Additionally, local building codes \citep[based on guidance such as][]{asce_24-05:2006,FEMA_p-55:2011}, may require all houses in flood zones to be elevated some margin above a nominal \gls{bfe}.
Design standards area also used to design large-scale infrastructure.
For example, some levees in the Netherlands are required to be designed such that the annual probability of failure is $<\nicefrac{1}{4000}$ years \citep{eijgenraam_flooding:2014}, while a seawall proposed as part of a \$29 billion coastal protection project proposed for  Galveston Bay, was designed for the annual probability of overtopping to equal 1\%, neglecting the effects of sea level rise \citep[Appendix D.,~p.~2-59]{USACE_coastal:2021}.

Risk-based design \citep{eijgenraam_flooding:2014,vandantzig_dike:1956}\ldots\james{Maybe add a bit here -- this is a low hanging critique}
And homeowners subject high flood risk can apply to programs that fund house elevation; decisions about which projects to fund are made primarily on the basis of cost-benefit ratio tests.
Like standards-based approaches, cost-benefit analyses or risk-based designs rely on probabilistic descriptions of relevant hazard.

Current methods for estimating the probability distributions that underpin cost-benefit analysis or design events focus on data-driven, objective methods that can be applied consistently across locations.
For example, \gls{usgs} Bulletin 17C specifies the procedures for estimating flood frequency and \gls{noaa} Atlas 14 provides estimates of the intensity duration and frequency of extreme rainfall.
A key assumption underlying these approaches is stationarity (see \cite{bulletin17c:2019}, p.~2, or \cite{atlas14_texas:2018} p.~A.4-42).
Yet changes in both global climate and local environmental conditions challenge this assumption in many cases \citep{Merz:2014gf,Milly:2008dg}.
For example, clear trends in extreme rainfall are apparent across much of southeastern Texas \citep{fagnant_spatiotemporal:2020,nielsen-gammon_txrainfall:2020}.
While methods have been proposed for incorporating trends into these analyses \citep[see][for a review]{Salas:2018ge}, these assume specific forms of the trend which may not represent the physical mechanisms that lead to nonstationarity or their uncertainty \citep{DossGollin:2019,Montanari:2014hl,Serinaldi:2015bq}.

\subsection{Emerging paradigms}

Many future hazards depend on intrinsically unpredictable human decisions (\eg, the rate of future greenhouse gas emissions) or on physical processes that are poorly constrained by existing data \citep[\eg, collapse of the West Antarctic ice sheet; see][]{deconto_antarctica:2016} and are therefore deeply uncertain \citep{keller_management:2021,walker_deep:2013,lempert_complex:2002}.
These deeply uncertain nonstationary hazards challenge not only the existing stationary estimates but, more fundamentally, the premise that objective estimates of future hazard exist and can be estimated empirically.

Recognizing the challenges of \gls{dmdu}, many frameworks for identifying robust decisions have been proposed.
Most emphasize the use of models in an exploratory (``what-if'') framework to learn about interactions between decisions and system dynamics \citep{bankes:1993}.
For example, robust decision methods \citep{lempert_shaping:2003} run models over large ensembles of possible futures to assess the performance of different policies, then apply statistical analysis to identify the conditions under which particular policies perform well or poorly.
These methods have been applied to coastal infrastructure planning by sampling \glspl{sow} from a \gls{pdf} of \gls{slr}, then identifying decisions that perform well over many possible futures \citep{sriver_sealevel:2018,lempert_slr:2012,wong_nola:2017}.
When many possible levers are available, and exhaustive sampling therefore infeasible, multiobjective robust decision methods add policy search (using optimization tools) to identify promising actions \citep{kasprzyk:2013,kasprzyk_denovo:2012,hadka_mordm:2015}.
In addition, many studies formally quantify robustness to deep unccertainties \citep{herman:2015,mcphail_robustness:2019} and use this as a criterion for policy comparison.

Although these methods for \gls{dmdu} have proven valuable in a wide range of settings, they still require necessarily subjective assessments about the likelihood of future conditions.
Even exploratory models require the analyst to choose which uncertainties are considered and how they are sampled.
When analyses integrate performance over many possible futures, whether to compute sensitivity of responses to inputs, estimate expectations of functions, or compute robustness metrics, they necessarily make assumptions about likelihood of different scenarios.
Although many analyses contend that specifying a joint probability distribution function over inputs is inappropriate under deep uncertainty, assumptions about the ranges and independence of parameters to sample are just as subjective as the choice of probability distribution \citep{schneider_scenarios:2002,quinn_exploratory:2020} and, indeed, can be interpreted  as a specific choice of probability distribution (we revisit this point in \cref{sec:analysis-condition}).
This motivates the development of decision analytic frameworks that draw from the strengths of \gls{dmdu} methods such as exploratory modeling, vulnerability assessment, robustness checks, and iterative stakeholder critique, but that embrace the necessity of modeling the likelihood of possible futures.

\subsection{Research gaps and objectives}

Recognizing that the assumptions underpinning any decision analysis will necessarily be subjective and imperfect, we draw inspiration from the philosophy of Bayesian model selection, particularly in the ``$\mathcal{M}$-open'' case where there is no true model to identify (\ie, all models are approximations of the truth).
This construct is a theoretical contrast to the ``$\mathcal{M}$-closed'' case, where the goal is to select the ``true'' model from a family of candidate models, and we find many parallels to deep uncertainty.
The theory of model selection in the $\mathcal{M}$-open case emphasizes a subjective view, both of probability itself and of the modeling process, fr which probability distributions offer a self-consistent language with which to reason about the unknown rather than a statement of objective truth (see \cite{gelman_philosophy:2013} or \cite{ramsey_probability:2016} for a philosophical discussion, \cite{jaynes_probability:2003} for a detailed reference, or \cite{Piironen:2017eh} for a discussion of methods for model selection).
From a practical perspective, this literature emphasizes the importance of iteratively building models, simulating the consequences of those models, and subsequently critiquing them \citep{gelman_workflow:2020}.
Since, by definition, the models considered in an $\mathcal{M}$-open world are not ``true,'' this iterative workflow aims to identify models that are ``useful'' \citep{Box:1976gu} and promote a dialogue amongst stakeholders and decisionmakers and analysts \citep{gelman_philosophy:2013}

In this paper we offer a first conceptual step towards bridging the fields of \gls{dmdu} and Bayesian model choice.
We consider the didactic case study of whether to elevate a hypothetical house, and if so how high, as a specific example of a decision problem where both over- and under-investment can lead to undesired outcomes.
Prior studies have found that floodproofing and building-scale vulnerability reduction measures, including house elevation, can effectively reduce local flood damages in many contexts \citep{demoel_reducing:2014,deruig_building:2020,kreibich_building:2005,slotter_floodproofing:2020,Rozer:2016dn,mobley_mitigation:2020,aerts_cost:2018}, and both local building codes \citep{asce_7-10:2013,bruneau_multihazard:2017,asce_24-05:2006} and federal policy \citep{FEMA_p-55:2011} require elevation in some cases.
Guidance for homeowners, notably from \gls{fema}, recommends elevating to the \gls{bfe} (typically the \SI{100}{year} flood) plus a freeboard \citep{fema_retrofitting:2014,asce_24-14:2015,fema_retrofitting:2014} but recent suggests scope for improvement.
For example, this ``one size fits all'' approach can deviate substantially from ``optimal'' guidance tailored to specific homes \citep{xian_elevation:2017}.
Further, that neglecting uncertainty in the cost-benefit analysis can lead to poor decisions \citep{zarekarizi_suboptimal:2020}.
Focusing on deep uncertainty in \gls{slr} over the 71 year design life of a hypothetical house, we seek to answer the research question ``\emph{how can deep uncertainties be transparently and consistently synthesized for decision analysis?}''

We proceed as follows.
In \cref{sec:analysis} we present three formal decision analytic frameworks for analyzing an ensemble of \gls{slr} simulations, building through existing approaches for exploratory modeling (\cref{sec:analysis-explore}) and scenario analysis (\cref{sec:analysis-explore}) towards a formal Bayesian method for transparently synthesizing deep uncertainty through subjective prior beliefs.
In \cref{sec:case-study} we describe the case study in detail.
Next in \cref{sec:results} we discuss results for each of the three theoretical lenses.
Finally in \cref{sec:conclusions} we discuss key findings, future research needs, and implications for research, policy, and practice.

\section{Decision analytic framework}\label{sec:analysis}

In this section we outline our framework for decision analysis under deep uncertainty, maintaining a high level of generality.
In the next we discuss application to the house elevation case study.

Following \cref{fig:flowchart}, consider using $J$ \glspl{sow} $\vb{s} = \qty{s_1, s_2, \ldots, s_J}$ (box b in \cref{fig:flowchart}) to evaluate $I$ discrete decisions $\vb{x} = \qty{x_1, x_2, \ldots, x_I}$ (box a in \cref{fig:flowchart}).
For each scenario $s_j \in \vb{s}$ and decision $x_i \in \vb{x}$ we use a system model $f$ to calculate a set of metrics describing the performance of decision $x_i$ on \gls{sow} $s_j$, which we denote $u_{ij} = f(x_i, s_j)$ (box c in \cref{fig:flowchart}).
While we assume for simplicity that the decision space is are known and finite, this approach could be coupled to a policy search model that proposes candidate decisions $x_i$ from some decision space $\mathcal{X}$.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{bayes-rdm.pdf}
    \caption{
        Outline of the proposed decision-analytic framework.
        In \cref{sec:analysis-explore} we use an exploratory framework to quantify the performance (c) of candidate decisions (a) under a large ensemble of possible futures (b).
        In \cref{sec:analysis-condition} we illustrate the ``multiple PDF problem'' by creating probability distributions over outcomes (e) that are conditional upon specific models describing the likelihood of different futures (d).
        In our case study, these are \gls{rcp} scenarios and physical models of sea level rise.
        Finally in \cref{sec:analysis-synthesize} we propose a subjective Bayesian framework for synthesizing across deep uncertainties by re-weighting sampled futures (f).
    }\label{fig:flowchart}
\end{figure}

\subsection{Explore}\label{sec:analysis-explore}

A first analytical step is to use the model in an ``exploratory'' mode.
Exploratory modeling avoids making explicit assumption about the likelihood different \glspl{sow} and instead seeks to generate new knowledge \citep{bankes:1993}, for example by discovering particularly relevant scenarios  \citep{lamontagne_discovery:2018,groves_scenarios:2007} and mapping a system's response to decisions and \glspl{sow} \citep{Poff:2015jn,Steinschneider:2015kk,sriver_sealevel:2018}.
Although exploratory modeling seeks to avoid explicit assumptions about the likelihood of different futures, subjective modeling decisions including the choice of system model, the set of candidate decisions, the criteria used to assess outcomes, and the choice of how to sample \glspl{sow} \citep[see][]{quinn_exploratory:2020} strongly influence results.
\klaus{Does this address your comment about linking back to observations?}

\subsection{Condition}\label{sec:analysis-condition}

Although exploratory modeling is a useful framework for understanding systems, there are many questions that it cannot answer.
For example, answering questions like ``what is the 95th percentile of some metric under decision $x$'' or ``which decision minimizes expected damages'' or ``what is the probability of failing to satisfy a critical metric''  requires estimating the expectation of some function $\pi$, which has the form
\begin{equation*}
    \mathbb{E}\qty[f(s | x_i)] = \int_{s \in \Omega} p(s) \pi(s | x_i) \dd{s}.
\end{equation*}
Because expectations are defined relative to a probability distribution, \emph{estimating any such expectation requires defining a probability distribution over \glspl{sow}}.

In boxes (d) and (e) of \cref{fig:flowchart} we add this concept to our decision analytic framework.
Given some model $M_k$ specifying a probability over \gls{sow} $p(\Omega)$, we sample $\vb{s} \sim p(\Omega | M)$.
Since the \glspl{sow} are drawn \gls{iid} from $M_k$, the set of outcomes $u_{i, j}$ can be interpreted as \gls{iid} draws from the conditional distribution over outcomes, $p(u | x_i, M_k)$.

Many decision analytic tools used in climate risk management can be interpreted through this lens.
A common application (to which we return in our case study) is when using probabilistic models with deeply uncertain boundary conditions.
For example, \ldots\james{Add a water resources and a flood example}
In this case, $M$ describes both the choice of model and the choice of boundary condition (\eg, choice of physical model and the \gls{rcp} scenario).
As we shall discuss in \cref{sec:results-conditional}, this analysis models uncertainty \emph{within models} probabilistically, but models uncertainty \emph{between models} only using a scenario approach (\ie, choices of model structures or factors such as forcings we show as cases) without any explicit probability statement \citep[see][]{wong_nola:2017}.\james{Wording}

Another application is to studies that generate an ensemble of \glspl{sow} by sampling parameters across fixed ranges.
For example, \citet{lempert_slr:2012} sample parameters describing the rate of \gls{slr} across a range of values to inform coastal adaptation.
Similarly, \citet{trindade_waterpathways:2020} checks the performance of candidate decisions against an ensemble of synthetic time series of streamflow, water demand, and other parameters by sampling parameters that transform the available data over a plausible range \citep[\ie, calculating robustness metrics; see][for details]{mcphail_robustness:2019,herman:2015}.
Since sampling over a range is equivalent to sampling from a Uniform distribution, this method makes an implicit assumption that $M$ is the product of $L$ independent Uniform distributions (one for each parameter) over the plausible ranges $\qty(\theta_{\ell,\min}, \theta_{\ell,\max}) \forall \ell \in \qty{1, \ldots, L}$:
\begin{equation}
    p(\Omega | M) = \prod_{\ell=1}^L \mathrm{Uniform} \qty(\theta_{\ell,\min}, \theta_{\ell,\max}).
\end{equation}
If this product of independent Uniform distributions truly represents a well-calibrated belief about the joint probability distribution over \glspl{sow} $p(\Omega)$ then it should be used, but analysts should be clear that, like it or not, they are making probabilistic assumptions, and should apply methods for model critique and validation \citep[see][]{gelman_workflow:2020}.

Our primary concern is not that subjective assumptions about the likelihood of different futures are wrong -- this is, almost surely, inevitable -- but that when these assumptions are opaque and presented without critique or validation they may lead to poor decisions.

\subsection{Synthesize}\label{sec:analysis-synthesize}

If the ensemble of \glspl{sow} available does not represent our subjective belief about the plausibility of future conditions, then some corrective measure should be taken.
One possibility is to discard the $\vb{s}$ and resample from some ``true'' distribution $p_\mathrm{belief}$.
In practice, this is impractical because simulating $\vb{s}$ may rely on complicated models that are computationally expensive to re-run.
Consequently, climate risk management often uses the outputs of structured simulations and analyses.
For example, many studies use outputs from \gls{cmip} simulations \citep{meehl_cmip:2000} to assess risks and design adaptation strategies.\james{Add some refs}
It is not practical or feasible for end users to re-run such complex Earth system models every time they want to sample \glspl{sow} from a different distribution.

Instead of attempting to re-sample $\vb{s}$, we instead consider re-weighting.
To do this we first project the \glspl{sow} $\vb{s} \in \Omega$ onto a low-dimensional representation, which we denote $\qty{\psi_1, \psi_2, \ldots, \psi_J} \in \Psi$.
This allows us to specify a prior belief over this reduced space $p_\mathrm{belief}(\Psi)$ instead of over the full space $p_\mathrm{belief}(\Omega)$.
We then calculate a probabilistic weight $w_j \in \qty[0, 1]$ for each \gls{sow} $s_j$ so that the weighted distribution of $\psi_j$ closely approximates $p_\mathrm{belief}(\Psi)$.

We present here the case where the $\psi_j$ are one-dimensional.
We first sort the $\psi_j$  from least to greatest so that $\psi_{j-1} \leq \psi_j$, ($j \neq 1$).
Defining $F_\mathrm{belief}(s)$ to be the cumulative distribution corresponding to $p_\mathrm{belief}$, we calculate weights as
\begin{equation}\label{eq:weights}
    w_j = \begin{cases}
        F_\mathrm{belief}\qty(\frac{1}{2}\qty[\psi_1 + \psi_2])                                                                     & j = 1     \\
        F_\mathrm{belief}\qty(\frac{1}{2}\qty[\psi_{j} + \psi_{j+1}]) - F_\mathrm{belief}\qty(\frac{1}{2}\qty[\psi_{j-1} + \psi_j]) & 1 < j < J \\
        1 - F_\mathrm{belief}\qty(\frac{1}{2}\qty[\psi_{J-1} + \psi_J])                                                             & j = J.
    \end{cases}
\end{equation}
This step is illustrated in \cref{fig:grid-sketch}.
For higher dimensional projections, this equation can be extended by partitioning the parameter space into a Voronoi diagram, and then integrating the \gls{pdf} $p_\mathrm{belief}$ over each region.
This approach is widely used in analysis of tropical cyclones: first an ensemble of cyclones is simulated, and then probabilistic weights are calculated for each based on attributes such as intensity, landfall location, and radius \citep{johnson_clara:2013,resio_probabilities:2007,toro_jpm-os:2010}.
From a practical perspective, many simulation models round outputs, meaning that there is a possibility of having $\psi_j$ that are equal; in this case it may be helpful to add a small amount of noise to the $\psi_j$ before computing the weights.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{grid-sketch}
    \caption{
        Schematic of \gls{sow} weighting scheme defined in \cref{eq:weights}.
        This method is illustrated for a hypothetical target distribution (orange line) and $J=4$ samples $\psi_1, \psi_2, \psi_3, \psi_4$ (orange dots).
        As shown in \cref{eq:weights}, the weights $w_j$ (vertical lines) are calculated based on the cumulative distribution function of the target distribution  (black line) at the halfway points $\frac{1}{2}\qty[\psi_i+\psi_{i+1}]$ (vertical dashed lines).
    }\label{fig:grid-sketch}
\end{figure}

The aim of this re-weighting framework is to integrate an ensemble of \glspl{sow} used for exploratory modeling into formal decision analysis.
As in \cref{sec:analysis-condition}, we must condition on a model: where the analysis of \cref{sec:analysis-condition} conditions upon deep uncertainties, the approach outlined in this subsection synthesizes across them.
It is crucially important to understand that stakeholders and experts will not, in general, agree on $p_\mathrm{belief}$ because its uncertainty is epistemic -- that is, there is no ``true'' value of $p_\mathrm{belief}$ that could be estimated if only we had better models and data.
However, reasoning using subjective probability allows\ldots\james{Finish sentence}

@@

We model a one-time decision of whether to elevate a house, and if so by how much (\cref{fig:xlrm}).
We focus on a case study of a \emph{hypothetical} house in Norfolk, VA.
For interpretability, we focus on deep uncertainty in \gls{msl} and treat other model parameters as more shallow uncertainties as shown in \cref{tab:uncertainties}.
We use the notation developed in the previous section to describe the case study.
Specifically,
\begin{enumerate}
    \item The decision vector $\vb{x}$ is a scalar describing how high to elevate a house ($\Delta h$); we consider $\Delta h = \qty{\SI{0}{ft}, \SI{0.25}{ft}, \ldots, \SI{12}{ft}}$ f
    \item The \glspl{sow} describe annual time series of \gls{msl} over the $T=71$ year house lifetime so $\vb{s} \in \mathbb{R}^T$
    \item The system model $f$ quantifies up-front costs and lifetime expected damages, given a decision $x_i$ and \gls{sow} $s_j$, by integrating economic and engineering damage models over a probability distribution for storm surge.
\end{enumerate}
In the remainder of this section we describe data sources and treatment of \gls{slr} (\cref{sec:case-slr}), storm surge (\cref{sec:case-surge}), damages and metrics (\cref{sec:case-metrics}), and finally the subjective priors $p_\mathrm{belief}$ used to apply the re-weighting method described in \cref{sec:analysis-synthesize} to this case study (\cref{sec:case-priors}).

>>>>

\documentclass[12pt]{article}

% Figures
\usepackage[demo]{graphicx}
\usepackage[list=true]{subcaption}
\graphicspath{{../../plots/}{../../tikz/}{../../img/}}
\usepackage[section]{placeins} % require floats to appear in the section they are defined

@@

% footer
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage[en-US]{datetime2}
\fancyhf{}
\fancyhead[L]{CONFIDENTIAL DRAFT by J. Doss-Gollin \& K. Keller}
\fancyhead[R]{\DTMnow}
\fancyfoot[R]{page~\thepage~of~\pageref{LastPage}}
\pagestyle{fancy}

@@

% authors
\usepackage{authblk}
\title{Mainstreaming Climate Adaptation Requires Subjective Integration of Deep Uncertainties}
\author[1]{James Doss-Gollin}
\author[2]{Klaus Keller}
\affil[1]{Department of Civil and Environmental Engineering, Rice University}
\affil[2]{Thayer School of Engineering, Dartmouth College}
\renewcommand*{\Affilfont}{\normalsize\normalfont}

% ACRONYMS
\usepackage[acronym,nopostdot,nonumberlist,shortcuts,]{glossaries}
\newacronym{bdt}{BDT}{Bayesian decision theory}
\newacronym{bfe}{BFE}{base flood elevation}
\newacronym[]{cdf}{CDF}{cumulative distribution function}
\newacronym{fema}{FEMA}{the Federal Emergency Management Agency}
\newacronym{gcm}{GCM}{general circulation model}
\newacronym[]{gev}{GEV}{generalized extreme value}
\newacronym{iid}{IID}{independent and identically distributed}
\newacronym{ipcc}{IPCC}{International Panel on Climate Change}
\newacronym{lsl}{LSL}{local mean sea level}
\newacronym{mcmc}{MCMC}{Markov Chain Monte Carlo}
\newacronym{pdf}{PDF}{probability density function}
\newacronym{rcp}{RCP}{representative concentration pathway}
\newacronym{msl}{MSL}{mean relative sea level}
\newacronym{rdm}{RDM}{robust decision making}
\newacronym[plural=SOWs,descriptionplural=states of the world]{sow}{SOW}{state of the world}
\newacronym{tc}{TC}{tropical cyclone}

@@

\begin{abstract}
    National agencies, regional governments, and individual households manage flood risk through a combination of standards and cost-benefit analyses that rely on estimates of flood hazard.
    One example is house elevation: \acrlong{fema} recommends a house located in the floodplain be elevated to the \acrlong{bfe} (typically the 100 year flood level) plus a freeboard (typically \SI{1}{ft}) if doing so would pass a cost-benefit test.
    Under stationarity, the probabilities used for this analysis (\ie to estimate the 100 year flood height or to perform the cost-benefit analysis) can be estimated using empirical frequencies from the observational record.
    However, growing recognition of nonstationarity motivates a forward-looking approach.
    Efforts to incorporate nonstationarity hazard into planning have struggled to overcome deep uncertainties in projections of future hazard because different models  may fit the observed data equally well but yield very different estimates of future hazard, and there is often no objective way to choose between them (\ie, equifinality).
    In this paper we use a didactic case study of deciding if and how high to elevate a single house in the coastal zone to illustrate a decision theoretic framework that combines exploratory modeling, robust decision making, and subjective Bayesian philosophy.
    We first use exploratory modeling to illustrate that there is no dominant strategy; instead, there are instrinsic tradeoffs between performance under low- and high-sea level rise scenarios.
    Next, we map the tradeoff between up-front cost and expected damages for each of sixteen models of future sea level rise to illustrate that the choice of which model to use has a strong influence on estimated trade-offs, cost-benefit analyses, or policy prescriptions.
    To overcome these shortcomings, we propose and illustrate a computationally efficient method for scenario weighting conditional on a subjective prior belief over future sea level rise.
    This method provides a transparent framework for synthesizing deep uncertainties, facilitating critique and dialogue around necessarily subjective decisions.
    It can be applied beyond nontationary flood risk management to other problems in climate adaptation, calculating the social cost of carbon, and more.
\end{abstract}

Key points
\begin{enumerate}
    \item
\end{enumerate}

@@

Estimates of flood hazard affect peoples' lives
\begin{enumerate}
    \item Floods affect lots of people and property
    \item Climate change and exposure making floods worse
    \item Floodplain management: $T$ year flood
    \item Some cost-benefit analysis and lots of design standards / heuristics
\end{enumerate}
One example is floodproofing
\begin{enumerate}
    \item Literature on floodproofing and building-scale vulnerability reduction measures
    \item This works great!
    \item But who should use it? We're still using standards.
    \item Prior studies have found that floodproofing and building-scale vulnerability reduction measures, including house elevation, can effectively reduce local flood damages in many contexts \citep{demoel_reducing:2014,deruig_building:2020,kreibich_building:2005,slotter_floodproofing:2020,Rozer:2016dn,mobley_mitigation:2020,aerts_cost:2018}.
          Official guidance for homeowners, notably from \gls{fema}, recommends elevating to the \gls{bfe} (typically the \SI{100}{year} flood) plus a freeboard \citep{fema_retrofitting:2014,asce_24-14:2015,fema_retrofitting:2014} but recent suggests scope for improvement.
          For example, \citet{xian_elevation:2017} used a cost-benefit analysis to demonstrate that tailoring recommendations to the inital elevation of a structure can reduce expected costs.
          \citet{zarekarizi_suboptimal:2020} show that neglecting uncertainty in discount rate, house lifespan, flood risk, and depth-damage curves can lead to maladaptation.
          Yet these studies are silent on the question of how nonstationary flood hazard should factor into this decision.
\end{enumerate}
Nonstationarity
\begin{enumerate}
    \item These analyses all rely on \glspl{pdf}
    \item Growing evidence for nonstationarity in a wide variety of hydroclimate processes
    \item We often focus on climate change as a driver but sometimes local modifications matter just as much \citep{Merz:2014gf}
    \item Even where risk-based analyses are used, this problem recommendations
    \item Yet stationarity remains the dominant assumption because it is immortal \citep{Montanari:2014hl} or undead \citep{Serinaldi:2015bq}
    \item Bonstationarity of flood hazard is already evident in many places, especially in coastal areas where \gls{msl} is clear.
    \item On the other, efforts to find objective methods for quantifying nonstationarity, eg of precipitation \citep{atlas14_texas:2018} or flood frequency \citep{bulletin17c:2019}, have struggled to find a satisfactory approach because factors like future emissions and localized climate response create deep uncertainties that resist objective quantification
\end{enumerate}
It matters how you prepare
\begin{enumerate}
    \item Equifinality problem: different models diverge even if they fit the historical record well
    \item Over- and under-design can each be costly \citep{DossGollin:2019}
    \item On the one hand, overly optimistic assumptions of future risk can lead to unacceptably high risk of failure in the future if additional investments are not made.
          For example, a failure to prepare Texas's interconnected natural gas, electricity, and water systems for temperatures that had been observed in the past 30 years caused cascading failures with disparate impacts \citep{doss-gollin_txtreme:2021,busby_cascadingrisks:2021}.
    \item On the other, unless opportunity costs are negligible, designing for a too-pessimistic scenario can contribute to maladptation by limiting resource availability in the future. For example,\ldots\ldots\james{add an example of a maladaptive megaproject -- look to sustainability science or the arguments in \citet{ansar_bigisfragile:2017}, something by Flyvbjerg, \etc}
\end{enumerate}
A quick philosophical literature review of how people have approached and interpreted this problem
\begin{enumerate}
    \item This motivates subjective approaches for designing for nonstationarity.
    \item There's a little bit of RDM-style work on floods \citep{sriver_sealevel:2018,garner_slrise:2018,lempert_slr:2012}
    \item Model choice in $\mathcal{M}$-open case
    \item Philosophy of Bayesian statistics and decision making -- explain subjective probability
\end{enumerate}
Case study
\begin{enumerate}
    \item House elevation in the coastal zone as a specific example of a problem where both over- or under-investment could be a problem
    \item Federal guidance uses both standards (floodplain, \gls{bfe}) and cost-benefit analysis (does it pass test?)
    \item Be sure to note that this has parallels to lots of other engineering design guidance.
    \item \citet{xian_elevation:2017}: one size does not fit all
    \item \citet{zarekarizi_suboptimal:2020}: neglecting uncertainty leads to bad outcomes
\end{enumerate}
Plan d'attaque
\begin{enumerate}
    \item Research questions
          \begin{description}
              \item[Q1] What are the potential benefits to homeowners of changing threshold-based guidance to risk-based guidance under nonstationary flood hazard?
              \item[Q2] How does risk-based guidance change under different assumptions of physical model structure and human emissions forcings?
              \item[Q3] How can deep uncertainties be transparently and consistently synthesized for decision analysis?
          \end{description}
    \item In \cref{sec:case-study} we describe our case study in Norfolk, VA.
    \item  in \cref{sec:multiple-simulation} we analyze each simulation (\ie, realizations from these \glspl{pdf}) independently and illustrate .
    \item  in \cref{sec:multiple-pdf} we consider the design problem as a set of ``multiple \glspl{pdf}'' and use these findings to motivate theoretical advances.
    \item  in \cref{sec:synthesizing} we present a probabilistic method for synthesizing simulations from multiple \glspl{pdf} and illustrate the advantages and potential pitfalls of this approach.
    \item  in \cref{sec:conclusions} we briefly summarize our findings, discuss implications for engineering practice, and identify  future research needs.
\end{enumerate}

@@

We model a single, one-time decision of whether to elevate a house, and if so by how much (\cref{fig:xlrm}), for a \emph{hypothetical} house in Norfolk, VA.
For interpretability, we focus on deep uncertainty in \gls{msl} and treat other model parameters as fixed quantities or as outputs from a single probabilsitic model, as shown in \cref{tab:uncertainties}.
We evaluate this decision for each of many possible simulations of \gls{msl} using a combined engineering-economic model based on that of \citet{zarekarizi_suboptimal:2020}.
In the remainder of this section we discuss data sources and models.

\begin{table}
    \centering
    \caption{
        Summary of parameters, their notation, and how their uncertainty is modeled.
    }\label{tab:uncertainties}
    \begin{tabular}{l l p{3in} l}
        \toprule
        Name             & Symbol            & Uncertainty                                                                          & See more                     \\
        \midrule
        \Gls{msl}        & $\overline{y}(t)$ & Deeply uncertain: four physical models $\times$ four \acrshort{rcp} scenarios        & \cref{sec:sea-level}         \\
        Storm surge      & $y'(t)$           & Probabilistic: Bayesian inference on a stationary \acrshort{gev} model               & \cref{sec:storm-surge}       \\
        Ann-max flood    & $y(t)$            & Deterministic: $y(t)=\overline{y}(t)+y'(t)$                                          &                              \\
        Discount rate    & $1-\gamma$        & Fixed at 3\%                                                                         & \cref{sec:led}               \\
        Depth-damage     & $D(h-y)$          & Deterministic: based on HAZUS model                                                  & \cref{fig:cost-depth-damage} \\
        Elevation cost   & $C(\Delta h)$     & Deterministic: a piecewise linear model following \citet{zarekarizi_suboptimal:2020} & \cref{fig:cost-up-front}     \\
        Initial height   & $h_0$             & Deterministic: \SI{1}{ft} below the \gls{bfe} unless otherwise noted                 &                              \\
        House floor area &                   & Deterministic: \SI{1500}{ft^2} unless otherwise noted                                &                              \\
        Structural value &                   & Deterministic: \usd{200000} unless otherwise noted                                   &                              \\
        House lifetime   &                   & Deterministic: 70 years unless otherwise noted                                       &                              \\
        \bottomrule
    \end{tabular}
\end{table}
